---
title: "tidyplate: A Versatile R Package for Microplate Data Transformation"
author: "Shubham Dutta"
output: word_document
date: "2024-11-04"
editor_options: 
  markdown: 
    wrap: 72
csl: mdpi.csl
bibliography: references.bib
---

## Abstract

Microtiter plates are critical tools in research and diagnostic
settings, enabling efficient, high-throughput analysis in assays like
ELISA, PCR, and immunoassays. However, working with plate-shaped data
present several challenges for data manipulation and analysis. While it
is easier to visualize plate-shaped data it is not ideal for downstream
data analysis and vice-versa. The *tidyplate* package addresses these
issues by transforming microplate-shaped data into tidy data back for
easy visualization and data analysis. *tidyplate* supports a variety of
standard plate formats and provides robust functions to validate,
convert, and export plate data. Here, we introduce the primary functions
of *tidyplate*, illustrate its applications, and discuss its benefits
for researchers.

## Introduction

Microtiter plates are standard tools in laboratories, offering
advantages in high-throughput applications, including tissue culture,
immunoassays, and diagnostic tests for infectious diseases such as HIV
and COVID-19. By allowing the handling of numerous samples in a compact
format, microplates streamline data collection and analysis, making them
highly efficient and versatile. Yet, the challenges associated with
transforming microplate data into analyzable formats remain significant.
On one hand it is easier to visualize plate-shaped data but on the other
hand it becomes difficult to perform data manipulation or data analysis.

Tidy dataframes are essential in data analysis because they offer a
standardized structure that makes data easier to understand and
manipulate [@tidy, @tibble]. In a tidy data, each variable forms a column,
each observation forms a row, and each type of observational unit forms
a table. This structure minimizes confusion and reduces errors in
analysis, as each element is clearly defined and easily accessible. Tidy
data facilitates interoperability between different data analysis tools
and packages, such as those in the R (Tidyverse) and python (Pandas),
which are designed to work seamlessly with tidy data [@tidyverse;
@mckinney-proc-scipy-2010]. By following a consistent format, tidy
dataframes enable analysts to write more efficient and readable code,
streamline workflows, and maintain reproducibility, making it easier to
share and scale analyses across projects and collaborators.

The *tidyplate* R package aims to bridge this gap, helping users
transforming various microplate layouts (6-well, 12-well, 24-well,
48-well, 96-well, 384-well, and 1536-well) from CSV or Excel files into
tidy data and back to original plate-shaped format (Figure
\@ref(fig:illus)). This enables researchers to view their data and when
necessary make them compatible with data manipulation and visualization
workflows in R and Python [@R; @CS-R9526]. In this article, we detail the key functions
of *tidyplate*, demonstrate usage examples, and discuss the practical
benefits for data-intensive applications.

## Usage and examples

*tidyplate* has two core functions: `tidy_plate` and `generate_plate`.
The `tidy_plate` function takes a specifically formatted input file (csv
or excel) as input and transforms it into tidy data (also known as a
dataframe or tibble in R). The data for an experiment must be stored
such that the top left corner should hold the name of the plate, the
column names for each plate should be named as 1, 2, 3 and so on and so
forth, the row names should be A, B, C, and finally there must be one
empty row between each plate. Figure \@ref(fig:tidy) shows how a csv or
excel file for a 12-well plate should be formatted and the same
principle can be used on other plate formats as well. In this experiment
researchers looked at the effect of antibiotics on cell lines. The
individual plates are named *drug*, *cell_line*, and *percent_survived*.
The row and column names are A, B, C and 1, 2, 3, 4 respectively. Each
plate is separated by an empty row. `tidy_plate` converts it into tidy
data by organizing each plate into columns and adding a *well* column
for the well identifiers. On the other hand `generate_plate` does the
opposite. It takes tidy data and reformats it back into plate-shaped
data. The fill colors are for illustrative purposes only.

Install *tidyplate*:

```{r}
#| label: install
#| eval: false
install.packages("tidyplate")
```

The users need to load the package using the `library` function and use
the `tidy_plate` function to convert an example plate into tidy data.

```{r}
#| label: tidy-demo
library(tidyplate)
file <- system.file("extdata", "example_12_well.xlsx", package = "tidyplate")
data <- tidy_plate(file)
head(data)
```

Any tidy data can be transformed back into plate shaped format using the
`generate_plate` function.

```{r}
#| label: gen-demo
#| eval: false
generate_plate(data, well_id = "well", plate_type = 12, file = "plate_shaped.csv")
```

In addition to the core functions, there are other helper functions to
help the users. One of the most useful of these functions is
`build_plate`. To assist with formatting, `build_plate` generates empty
csv or xlsx templates for each plate type, ensuring compatibility with
`tidy_plate`.

```{r}
#| label: build-demo
#| eval: false
build_plate(plate_type = 96, n_plates = 2, file = "template_96.xlsx")
```

If the users have formatted their own input files they can use the
`check_plate` function to ensure correct formatting before data
processing. `check_plate` verifies if a file adheres to *tidyplate*'s
input requirements. This function detects formatting issues, such as
incorrect row and column identifiers, ensuring smoother workflows.

```{r}
#| label: check-demo
#| error: true
check_plate(file) # No error for valid file

invalid_file <- system.file("extdata", "incorrect_format.csv", package = "tidyplate")
check_plate(invalid_file) # Error type displayed
```

Finally, an utility function called `view_plate_names` retrieves the
names of individual plates within a file, providing a quick reference
for users working with multi-plate files.

```{r}
#| label: view-demo
view_plate_names(file)
```

## Conclusion

*tidyplate* is a powerful tool for researchers and analysts, simplifying
the transformation of microplate data into tidy formats compatible with
R and Python [@R; @CS-R9526]. By addressing common challenges in plate data handling,
*tidyplate* supports a wide range of research applications, from
diagnostics to high-throughput screening. With its flexible functions,
researchers can validate, format, and analyze microplate data
efficiently, improving the reproducibility and accessibility of their
workflows.

## References

::: {#refs}
:::

\newpage

## Figures

Figure 1

![tidyplate transforms microwell plate-format data to tidy data and
back](paper_fig/tidyplate_illus.png){#fig:illus}

\newpage

Figure 2

![An illustration of how the core *tidyplate* functions `tidy_plate` and
`generate_plate` work on a 12-well
plate](paper_fig/tidy_plate.png){#fig:tidy}

\newpage

## Figure details

**Figure 1**: tidyplate transforms microwell plate-format data to tidy
data and back for visualization, data manipulation, and analysis.

**Figure 2**: An illustration of how the core *tidyplate* functions
`tidy_plate` and `generate_plate` work on a 12-well plate.
